Computational Methods for Principle Component Analysis Applied to Genotype Data

### 2018-05-01 doc/paper.tex
Finish addressing all Shawn's editings

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[semicolon,authoryear]{natbib} \bibliographystyle{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}
\usepackage{float}

@@

Population stratification (PS) is an important confounder for the association  between
genotypes and phenotypes (\cite{price2006principal}).
To adjust for PS, we need to accurately estimate the ancestry structure in the study samples. Principle component analysis (PCA) is a multivariate statistical method which finds the direction of the maximum variability. By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure  (\cite{reich}). To adjust for PS, PCA can be applied to study data to calculate PC scores, which are regarded as variables of ancestry direction and can be used as covariates to adjust for. An alternative approach is predicting the PC scores of the study samples using reference genotyped samples with detailed ancestry information.
This prediction-based approach not only allows to adjust for PS but also to infer ancestry memberships of study samples.

The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
We call this method simple projection (SP).
However, when the number of features are greatly larger than the size of the reference sample,
the predicted PC scores by this method are known to be biased toward null.
One way of addressing this shrinkage bias is presented by \cite{wang}.
Their solution is to combine one study individual with the reference set and
find the PC scores of this augmented data set by singular value decomposition (SVD).
The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
We call this method ``augment, decompose, and Procrustes transform'' (ADP).
This method has been shown to be effective in eliminating the shrinking bias of
study PC scores.
However since ADP needs to run SVD for each study samples, ADP is computationally expensive.

To address the limitations of SP and ADP,
we propose and evaluate two alternative PC score prediction methods.
The first approach removes the bias in SP by estimating the asymptotic bias factor,
which is calculated based on Random Matrix Theory \cite{dey}.
The second approach improves the computation efficiency of ADP by using online SVD algorithm. 
We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and and Procrustes transform'' (OADP).

@@

\subsubsection{Principle Component Analysis}\label{pca-intro}
Given a multi-dimensional data, PCA searches for the directions in which the data has the greatest variance.
These unit vectors are called the principle components (PCs).
The PC's form an orthonormal basis for the sample space,
and we can ignore all but the first few PC's to reduce the dimension of the data.
In particular, let $X=UDV^T$ be the SVD of $X$,
where $U$ is a $p \times n$ matrix whose columns are the PC's,
$D$ is an $n \times n$ diagonal matrix storing the standard deviations of the PC's,
and $V$ an $n \times n$ matrix such that each row contains the standardized PC scores of an individual.
(Alternatively, we can use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
We sort the diagonal elements of $D$ in descending order and rearrange the
columns of $U$ and $V$ accordingly.
We are interested in the top $k$ PC's (where $k \ll n$) and look at the top $k$
columns of $V$ for their PC scores.

@@

We assume that the reference data and the study data are generated from the same
population.
When the genotypes of all individuals of the population are given (which is very
unlikely in global genetic studies),
we use the method in \Cref{pca-intro} to find the PC scores of the individuals
we are interested in.
In practice, we have a reference dataset and a study dataset assumed to be generated from the
same population and use the former to estimate the populational PC scores of the latter.
There are several computational methods one can use to do this estimation.
In \Cref{sec:compu}, we compare the performance of four of them.

@@

\subsubsection{Simple Projection (SP)}
The standard way for estimating PC scores is by projecting the study
individuals' data points to the PC space of the reference individuals by using reference sample PC loadings.
We call this method ``simple projection'' (SP).
The algorithm goes as follows.

@@

As the reference sample size increases,
the computational complexity of SP stays as a constant,
except for decomposing the reference matrix, which is done once for all study individuals.
However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$,
which we call the ``shrinking bias''. (See \cite{dey}.)
This limits the accuracy of SP for high dimensional data.

@@

One way to solve the shrinking bias of SP is the ADP approach developed by \cite{wang}.
In this method, each study individual's data point is appended to the reference
data matrix.
Then SVD is applied to the augmented matrix,
and PC scores in this augmented sample are found for both the reference
individuals and the study individual.
The the PC scores of the reference individuals in this augmented sample is
compared with the PC scores of the reference individuals in the reference
sample,
which is calculated in the beginning,
and a linear (Procrustes) map is constructed to transform the latter into the former
with the smallest error.
This transformation is then applied to the PC scores of the study individual in
the augmented sample.
The resulted PC scores are used as an estimate for the study individual's
population PC scores.
Specifically, let
\[
  V =
  \begin{bmatrix}
    V_{\text{ref}} & W
  \end{bmatrix},
  \quad
  \tilde{V} =
  \begin{bmatrix}
    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
  \end{bmatrix},
\]
where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
Next, we use Procrustes analysis to find
\[
  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
for some metric $d$.
Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
The algorithm is summarized as follows:
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
  where $k_1 < k_2 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next study individual.
\end{enumerate}

@@

Although accurate, ADP is the slowest among the other three computational methods.
Its computational complexity increases cubically as the reference sample size and linearly as the study sampl sizes.
The ADP method is has been implemented in the TRACE software by \cite{wang}.

@@

A more efficient way to solve the shrinking bias of SP is to estimate the
magnitude and angle of the shrinkage and adjust the study individuals' PC scores accordingly.
This method has been developed by \cite{dey} and implemented in the High
Dimension Principal Component Analysis (HDPCA) in the R language.
HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias. ($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
\end{enumerate}

@@

In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
computation and memory cost of PCA on the reference samples can be enormous.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
When the reference sample size is small, RSVD can be slower than standard SVD,
but it is scalable to large sample size.
For the details of the RSVD algorithm, see \cite{halko} and \cite{abraham}.

@@

\subsection{Data}

@@

\bibliographystyle{plain}
\bibliography{main}

>>>>

\usepackage{epsfig} \usepackage[margin=1.5in]{geometry}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cleveref}
\usepackage{verbatim}
\usepackage{float}

@@

Population stratification (PS) is an important confounder for the association  between
genotypes and phenotypes \citep{price2006principal}.
To adjust for PS, we need to accurately estimate the ancestry structure in the study samples. Principle component analysis (PCA) is a multivariate statistical method which finds the direction of the maximum variability. By aggregating information across all genetic markers, PCA can accurate infer the ancestry structure \citep{reich2008principal}.
To adjust for PS, PCA can be applied to study data to calculate PC scores, which are regarded as variables of ancestry direction and can be used as covariates to adjust for. An alternative approach is predicting the PC scores of the study samples using reference genotyped samples with detailed ancestry information.
This prediction-based approach not only allows to adjust for PS but also to infer ancestry memberships of study samples.

The standard way of predicting PC scores is to find the PC loadings of the reference data and project the data points of the study samples onto them.
We call this method simple projection (SP).
However, when the number of features are greatly larger than the size of the reference sample,
the predicted PC scores by this method are known to be biased toward null \citep{dey2016asymptotic}.
One way of addressing this shrinkage bias is presented by \citet{wang2014ancestry, wang2015improved}.
Their solution is to combine one study individual with the reference set and
find the PC scores of this augmented data set by singular value decomposition (SVD).
The PC scores of the study individuals are then retrieved from the singular vector matrices and mapped to the reference PC space by a Procrustes transformation.
We call this method ``augment, decompose, and Procrustes transform'' (ADP).
This method has been shown to be effective in eliminating the shrinking bias of
study PC scores.
However since ADP needs to run SVD for each study samples, ADP is computationally expensive.

To address the limitations of SP and ADP,
we propose and evaluate two alternative PC score prediction methods.
The first approach removes the bias in SP by estimating the asymptotic bias factor,
which is calculated based on Random Matrix Theory \citep{dey2016asymptotic}.
The second approach improves the computation efficiency of ADP by using online SVD algorithm \citep{halko2011finding}. 
We call the first approach ``adjusted projection'' (AP), and the second approach ``online augment, decompose, and Procrustes transform'' (OADP).

@@

\subsubsection{Principle Component Analysis}\label{pca-intro}
Given a multi-dimensional data, PCA searches for the directions in which the data has the greatest variance.
These unit vectors are called the principle components (PCs).
The PC's form an orthonormal basis for the sample space,
and we can ignore all but the first few PC's to reduce the dimension of the data.
In particular, let $X$ be standardized so that each row (feature) has mean equal to zero and standard deviation equal to one.
Moreover, Let $X=UDV^T$ be the SVD of $X$,
where $U$ is a $p \times n$ matrix whose columns are the PC's,
$D$ is an $n \times n$ diagonal matrix storing the standard deviations of the PC's,
and $V$ an $n \times n$ matrix such that each row contains the standardized PC scores of an individual.
(Alternatively, we can use eigendecomposition to find the matrices since $X^T X = VD^2V^T$ and $X X^T = UD^2U^T$.)
We sort the diagonal elements of $D$ in descending order and rearrange the
columns of $U$ and $V$ accordingly.
We are interested in the top $k$ PC's (where $k \ll n$) and look at the top $k$
columns of $V$ for their PC scores.

@@

population.
In the ideal case in which the genotypes of all individuals of the population are given
we use the method in \Cref{pca-intro} to find the PC scores of the individuals
we are interested in.
In practice, however, we have a reference dataset and a study dataset assumed to be generated from the
same population and use the former to estimate the populational PC scores of the latter.
There are several computational methods one can use to do this estimation.
In \Cref{sec:compu}, we compare the performance of four of them.

@@

\subsubsection{Simple Projection (SP)}
The standard way for estimating PC scores is by projecting the study
individuals' data points to the PC space of the reference individuals by using reference sample PC loadings,
which is the matrix $U$ in the eigendecomposition of the (standardized) reference sample matrix $X$.
The PC loadings are unit vectors pointing toward the directions of the greatest variances in the $p$-dimension sample space.
We call this method ``simple projection'' (SP).
The algorithm goes as follows.

@@

As the reference sample size increases,
the computational complexity of SP stays as a constant,
except for decomposing the reference matrix, which is done once for all study individuals.
However, it is known that the projection method tends to underestimates the magnitude of the PC scores when $n \ll p$ \citep{dey2016asymptotic},
which we call the ``shrinking bias''.
This limits the accuracy of SP for high dimensional data.

@@

One way to solve the shrinking bias of SP is the ADP approach developed by \citet{wang2014ancestry, wang2015improved}.
In this method, each study individual's data point is appended to the reference
data matrix.
Then SVD is applied to the augmented matrix,
and PC scores in this augmented sample are found for both the reference
individuals and the study individual.
The the PC scores of the reference individuals in this augmented sample is
compared with the PC scores of the reference individuals in the reference
sample,
which is calculated in the beginning,
and a linear (Procrustes) map is constructed to transform the latter into the former
with the smallest error.
This transformation is then applied to the PC scores of the study individual in
the augmented sample.
The resulted PC scores are used as an estimate for the study individual's
population PC scores.
Specifically, let
\[
  V =
  \begin{bmatrix}
    V_{\text{ref}} & W
  \end{bmatrix},
  \quad
  \tilde{V} =
  \begin{bmatrix}
    \tilde{V}_{\text{ref}} & \tilde{W}_\text{ref} \\
    \tilde{V}_\text{stu} & \tilde{W}_\text{stu}
  \end{bmatrix},
\]
where the dimensions of $V_\text{ref}$, $\tilde{V}_\text{ref}$, and $\tilde{V}_\text{stu}$ are $n \times k_1$, $n \times k_2$, and $1 \times k_2$, respectively, with $ 1 \leq k_1 \leq k_2 \leq n+1$. 
Next, we use Procrustes analysis to find
\[
  A, b = \argmin{A, b} d(V_\text{ref}, \tilde{V}_{\text{ref}}A+b)
\]
for some metric $d$.
Finally, we get the Procrustes-adjusted top $k_1$ PC scores for the study individual $V_\text{pro} = \tilde{V}_\text{stu} A + b$.
The algorithm is summarized as follows:
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$ (columns of $D$ and $V$ are sorted by PC variance with the first column corresponding to the greatest PC variance).
  ($\bO[n^3]$.)
\item Append $X^T y$ and its transpose to the last column and row of $X^T X$, respectively.
  The entry of the bottom-right corner is $y^T y$.
  This is exactly the same as computing $\tilde{X}^T \tilde{X}$,
  where $\tilde{X} = (X, y)$ is the reference group with the study individual added.
  ($\bO[np]$.)
\item Apply eigendecomposition to $\tilde{X}^T \tilde{X}$ to get $\tilde{X}^T \tilde{X} = \tilde{V} \tilde{D}^2 \tilde{V}^T$.
  ($\bO[n^3]$.)
\item Project the first $k_2$ columns of $\tilde{V}$ to the first $k_1$ columns of $V$ by using Procrustes analysis to get the transformed PC scores of the study individual (the last row of $V$),
  where $k_1 < k_2 \ll n$.
  (Computational complexity is ignorable if $k_2$ is not large compared to $n$.)
  \item Go to step 3 for the next study individual.
\end{enumerate}

@@

Although accurate, ADP is the slowest among the other three computational methods.
Its computational complexity increases cubically with respect to the reference sample size and linearly with respect to the study sample size.
The ADP method is has been implemented in the TRACE software \citep{wang2014ancestry, wang2015improved}.

@@

A more efficient way to solve the shrinking bias of SP is to estimate the
magnitude and angle of the shrinkage and adjust the study individuals' PC scores accordingly.
This method has been developed by \citet{dey2016asymptotic} and implemented in the High
Dimension Principal Component Analysis (HDPCA) in the R language \citep{hdpca}.
HDPCA reads the PC scores and variances calculated by projection and estimate the shrinkage factor to restore the unshrinked scores. 
The algorithm is
\begin{enumerate}
\item Compute $X^T X$.
  (Computational complexity: $\bO[n^2p]$.)  
\item Apply eigendecomposition to $X^T X$ to get $X^T X = V D^2 V^T$.
  ($\bO[n^3]$.)
\item Calculate $W = Y^T (X V D_1^{-1})$, where $Y$ is the data matrix for all the study individuals, $D_1$ the first $k_1$ columns of $D$. ($\bO[mpk_1 + pnk_1]$.)
  \item Input $D$ and $W$ to the pc\_adjust function in the HDPCA function to obtain the PC scores adjusted for the shrinking bias. ($\bO[p]$ for finding the number of distant spikes (function select.nspike), $\bO[n]$ for adjusting for the shrinking bias (function hdpc\_est))
\end{enumerate}

@@

In addition to estimating the PC scores of the study individuals,
we have another algorithm that aims at improving the decomposition of the
training data.
For SP, AP, ADP, and OADP, we need to find the covariance matrix of the training
data, which has dimension $n \times n$, and conduct eigendecomosition on it.
However, when $n$ is exceedingly large,
computation and memory cost of PCA on the reference samples can be enormous.
A randomized SVD (RSVD) algorithm has been developed to handle this issue.
When the reference sample size is small, RSVD can be slower than standard SVD,
but it is scalable to large sample size.
For the details of the RSVD algorithm, see \citet{halko2011finding}.

@@

\section{Data}

@@

\bibliographystyle{plainnat}
\bibliography{frugalpca}
